---
title: "Case Study: Boston Housing Dataset"
author: "Emma Bennett, Pallavi, Lily"
date: 'UC3M, 2025'
output:
  html_document: 
    css: my-theme.css
    theme: cerulean
    highlight: tango
    number_sections: no
    toc: no
    toc_depth: 1
  pdf_document:
    css: my-theme.css
    theme: cerulean
    highlight: tango
    number_sections: yes
    toc: yes
    toc_depth: 1
editor_options:
  chunk_output_type: console
---
```{r}
library(mlbench)
library(tidyverse)
```

## Information about dataset
Found this dataset here: https://search.r-project.org/CRAN/refmans/mlbench/html/BostonHousing.html

According to the description it consists of housing data for 506 census tracts of Boston from the 1970 census. 

This dataset that we chose is good for a lot of these dimensionality reduction techniques (like factor analysis and PCA) because it has a good number of quantitative variables and is well-structured and easy to visualize.

The following are the columns of the dataset:

- **CRIM** — per capita crime rate by town  
- **ZN** — proportion of residential land zoned for lots over 25,000 sq.ft.  
- **INDUS** — proportion of non-retail business acres per town  
- **CHAS** — Charles River dummy variable (1 if tract bounds river; 0 otherwise)  
- **NOX** — nitric oxides concentration (parts per 10 million)  
- **RM** — average number of rooms per dwelling  
- **AGE** — proportion of owner-occupied units built prior to 1940  
- **DIS** — weighted distances to five Boston employment centres  
- **RAD** — index of accessibility to radial highways  
- **TAX** — full-value property-tax rate per \$10,000  
- **PTRATIO** — pupil-teacher ratio by town  
- **B** — 1000(Bk − 0.63)² where Bk is the proportion of blacks by town  
- **LSTAT** — % lower status of the population  
- **MEDV** — median value of owner-occupied homes in \$1000’s



## Loading data

```{r}
data("BostonHousing", package = "mlbench")
summary(BostonHousing)
```

```{r}
glimpse(BostonHousing)
```

## Data preprocessing
Below we are going to do some preprocessing on the data including dealing with outliers, checking if there are any NA values, and other feature engineering.

### Outliers

### NA/Missing Values
In this dataset there are no missing values.
```{r}
barplot(colMeans(is.na(BostonHousing)), las=2)
```

To show how we would deal with missing values going to remove NA values on a different dataset.


### Feature Engineering


## Visualization Insights

## Regression models and resampling tools

## PCA

## Factor Analysis

## Clustering tools

